Combination of field and experimental data together with computational models reveal cognitive mechanisms behind cleaner fish behaviour

### 2022-02-28 manuscript_1.0/manuscript_1.0.tex
New version of the manuscript. Using figure from the old data.

Disentangling the mechanistic underpinnings of behaviour
is typically done under controlled conditions in a laboratory setting. However,
controlled laboratory conditions often mask the inter and intra-specific
variation that arises in the interaction between mechanisms and
environmental variation. From a evolutionary perspective, mechanisms
are likely selected because of how they allow individuals to respond to
environmental variation. For example, biological market theory predicts that
the exchange rate of goods and/or services traded between cooperative
partners adjusts to the law of supply and demand, when individuals have
some degree of partner choice \citep{noe_Biological_1995a}.
Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Client fish seek cleaner fish services at their territory
(so-called ``cleaning station'') and offer themselves as food patches
to get their ectoparasites removed, which provides cleaners
with food and clients with improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016, demaire_Reduced_2020}.
Given the capacity of some client fish to swim larger distances and
access multiple cleaning stations while others access the only cleaning
station at their territory, it is crucial to categorize clients as
either ``visitors'' or ``residents'', respectively. During cleaning interactions,
a cleaner fish often face a choice between a visitor and a resident client
seeking its cleaning services simultaneously. Visitors have the option
to switch to another cleaner fish if being made to wait, while residents
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}.

When researchers aimed at testing wild-caught cleaner fish, as
well as individuals from other species, abilities
to prioritize visitor over resident clients in a lab-based paradigm,
they used Plexiglas plates of different colours and shapes offering same
amount of food as surrogates for visitor and resident clients.
One plate is made to behave like a resident, i.e.~it remains until
the cleaner fish had eaten off it. The other plate was made to behave
like a visitor, i.e.~it is removed if not inspected first.
Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Furthermore, adult cleaner fish outperformed various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; yet the African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. Given that primates
are known to readily distinguish
between one and two reward units if presented in more conventional ways
\(\color{red}{\text{ref}}\), this raises the question:
what makes the task difficult for most non-cleaners?

@@

To solve the ephemeral reward task, cleaners need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is in principle
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, the two mechanisms should
be evaluated in light of how well they explain the available data.

@@

The interplay between the underling cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether operant conditioning, chaining, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show low preference for visitors under low
local cleaner-to-client ratio.

@@

The model shows that agents need to find a mean
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates. If
\(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action.
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor as a negative component. Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The cognitive performance data was from a total of 120 cleaners
(10 individuals per 12 site/year) tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner fish inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. In the original studies, once a fish reach a learning criterion,
that is, performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)),
they passed to a reversal version of the task where the roles of the
visitor/resident Plexiglas plates were swapped (see Triki et al.~2019, 2020).
Here, we used instead a subset of these data in order to have an idea on
cleaner fish preferences for the visitor plate, even if they do not reach
the learning criterion within 200 trials. To do so, we first extracted the
trial-by-trial outcomes from the last two session (20 trials) of those who
never reached the learning criterion for visitor plate
(\(N = 45\) cleaner fish).
For those who reached the learning criterion at some point during the
test and passed to a reversal phase, we extracted the trials-by-trial
outcomes from the last session (10 trials) before passing to reversal
and the last 10 trials they were exposed to in the test
(\(N = 75\) cleaner fish).

@@

Figure \ref{fig:post} panels a,b and c show the posterior distributions
of the parameter values of the full model, which includes both chaining
and operant conditioning (negative reward). Posterior distributions show the probability
density (y axis) of the estimated parameter values (x axis). Notably,
the bulk of the posterior distribution for the parameter tuning negative
reward (\(\eta\) panel b) was around zero. Thus, the shape of the posterior
and its credible intervals suggest the absence of negative reward to be
well supported by the data. As for \(\gamma\), the confidence intervals
also includes zero, but the mode of the posterior is around 0.5. Thus,
we ran the analysis setting \(\eta\) to zero. Figure \ref{fig:post} panels
d and e, show the posterior distributions of the two parameters
estimated. Not surprisingly, under a model without negative reward, the
distribution of \(\gamma\) shifted to higher values. Under the new model,
the estimates of \(\gamma\) did not include zero and the distribution was
centred around \(0.5\). In panel f of figure \ref{fig:post}
we show the distribution of \(pseudo-R^2\) calculated using
samples from the posterior distributions shown before,
as well as from a model with negative reward but without
future reward. Note, \(pseudo-R^2\) can have negative values, which is when
the log-likelihood of the model is lower than that of a model that
triggers neutral preferences. Even though, the peak of the three
\(pseudo-R^2\) distributions was not very different, the model with only
future reward produced a distribution of \(pseudo-R^2\) where more values
were positive (to the right of black line in Fig. \ref{fig:post} f). This
shows that, accounting for variation in the parameter estimates, the
model with only future reward gives a better fit to the data, despite
having one parameter less.

Figure \ref{fig:pred} shows predictions from the two models,
one including chaining and operant conditioning (upper panels) and
one with only chaining (lower panels). On the
left, the colour coded contour shows the prediction of the model for
combinations of cleaner fish abundance (x axis) and visitor leaving
probability (y axis). Predictions were generated by running the model
using the mode of the posterior distributions as the parameter values,
and varying systematically cleaner fish abundance and visitor leaving
probability. Colour of the dots on top of the contour show the average
frequency with which cleaner fish chose the visitor plate in their last
20 experimental trials. Each dot corresponds to a reef-site/year of
data collection with a combination of cleaner fish abundance
and visitor leaving probability. Panels
on the right show how close the observed values for the probability of
choosing the visitor (y axis) were from the predictions of the models (x
axis). The black line represents the perfect fit between observed and
predicted values. Bars show the variation in predicted values using 100
samples from the posterior distribution. Despite having one more
parameter varying, the full model
gave a lower fit to the data than the model without negative reward.
In the supplementary material, we also show predictions from
model with only negative reward, excluding future rewards (\(\gamma=0\),
Fig. \ref{fig:nogamma}), provides a lower fit than either of the ones
presented in the main text.

The main reason for chaining and operant conditioning to give different
predictions is the way that cleaner fish relative abundance influences the
preference for the visitor clients. Visitor leaving probability has a
similar positive effect on the probability of choosing the visitor
clients on all three models (Fig. \ref{fig:pred}, \ref{fig:nogamma}). In
contrast, cleaner fish relative abundance has a different effect in the
model with only chaining, compared to the other two models,
full model and only operant conditioning. In the model
with only chaining, only intermediate cleaner fish abundance triggers a
preference for the visitor clients (Fig. \ref{fig:pred} c). In models
with negative reward, both intermediate and low cleaner abundance
triggered a preference for the visitor. (Fig. \ref{fig:pred} a,
\ref{fig:nogamma}).

@@

Our results support chaining as the more plausible mechanism behind the
intra-specific variation in cleaners' performance. This suggest
that chaining could be under positive selection. In the standard ephemeral
reward task, however, food amount and quality are equal in the
two Plexiglas plates. Since our computational models simulate this aspect of
the task, choosing visitors over residents at low cleaner fish abundances
does not diminish the amount of food/reward in the models.
On the other hand, a cleaner fish committed to visitors might find it more difficult
to include in its decision making other variables, like parasite load
and/or mucus quality \citep{roche_Client_2021}. Indeed, in the absence of
market effects, both field observations and laboratory experiments also
using simultaneous two choice options, show that cleaner fish base their
decision-making on other cues such as the mucus quality and the amount of food.\\
\citep{triki_Fluctuations_2019, grutter_Cleaner_2004}, as well as client size
as a proxy for food quantity \citep{wismer_Cuebased_2019, grutter_Does_2005}.
How important this limitation is will determine whether chaining yields a higher
fitness for cleaners.

Previous research showed that cleaner fish living at high population
densities and giving service priority to the visitor plate in the
ephemeral reward task, as well as cleaner fish living at low
densities but denying service priority to the visitor plate possess
larger forebrains; a key teleost brain region associated with
behavioural flexibility and social intelligence. Those failing to
show optimized decision-rules given their local ecological
conditions had relatively smaller forebrains \citep{triki_Brain_2020}.
Triki \emph{et al.} refer to the former as socially competent cleaner fish,
while the second group as socially incompetent cleaner fish.
Social competence is the ability to optimise social behaviour
to the available social information \citep{taborsky_Social_2012a, bshary_Cooperation_2015, varela_Correlated_2020}. Our analyses yielded no
evidence that the difference in social competence with respect to the local
ecological conditions and associated brain morphology,
found by Triki \emph{et al} \citep{triki_Brain_2020}, are due to the mechanism
used to incorporate future consequences. It could have been conceivable that
high performing individuals from low population densities reef sites
might use negative reinforcement instead of chaining, but in that
case negative reinforcement should have explained at least
part of the data. Configurational learning or
chunking \citep{sutherland_Configural_1989, miller_Magical_1956},
the second component necessary to solve the ephemeral
reward task \citep{quinones_Reinforcement_2019}, was
not varied in the models we analized here. However, while chunking tendencies
should vary to allow individuals to adapt to their local conditions
\citep{prat_Modelling_2022, kolodny_Evolution_2014}, systematic differences
in individual chunking tendencies would not explain how socially
competent decisions vary as a function of relative abundance.
Therefore, it remains currently unclear what cell-demanding
mechanisms may cause variation in social competence that
translates into site-specific variation in performance in the
ephemeral reward task.

Our models are inspired by the general processes of associative learning
where short term rewards are translated into decision making; thus, it
ignores alternative channels of information that could be relevant in
market-like situations. For example, the model does not investigate whether
cleaner fish actually assess the frequency of client visits or a mean frequency
of visitors leaving. The updating learning mechanism for the development
of preferences works on a trial-by-trial basis. In the model, cleaner fish
do not need to assess the actual state of the market, \emph{i.e.} their abundance,
the abundance of residents and visitors, and client visitation rate as an
indicator of demand. They only need to assess short-term consequences of their
own decisions on food intake, and chain them. Also, for the sake of
simplicity, the model ignores the process by which cleaner fish discriminate
residents and visitor clients. A model that accounts for this discrimination
probably would involve the development of preferences for morphological or
behavioural features that are statistically associated with visitors
or residents. For example, visitors are on average larger
than residents in body size \citep{bshary_Cleaner_2001a}, and contrary to
resindents, are less likely to chase a cleaner fish that fails to cooperate
and intead cheats its client by taking a bite of mucus
\citep{bshary_Asymmetric_2002}. Given these associations, chaining might
produce the decision-rule ``choose the larger client and/or the less aggressive
client'', which is not a useful rule in the standard ephemeral reward task.

>>>>

Often alternative cognitive mechanisms yield similar behaviour and/or cognitive
performances. This poses a problem for disentangling the mechanistic
underpinnings of behaviour. This is particularly clear in research aimed at
discovering interspecific variation in \emph{higher} cognitive abilities; or
in other words, research on whether non-human animals show cognitive
abilities believed to be uniquely human. For instance, when researchers
try to find \emph{mental time travel-like} behaviour, they usually come up with
clever experiments showing the behaviour displayed requires inferences made
through past events\citep{dally_Foodcaching_2006}. However, they often face challenges proposing alternative scenarios where simple associations can bring about the
predicted behavioural outcome \citep{suddendorf_Evolution_2007}. Similarly, attempts
to show \emph{theory of mind} in non-human animals face objections
justified by alternative mechanisms that bring about similar behavioural
results \citep{heyes_Theory_1998}. Such controversies are usually settled by
using the principle of parsimony, and it's cognitive version,
Lloyd Morgan's cannon. Which states that the simpler explanation (mechanism)
must be accepted. Ideally, however, alternative hypothesis should be evaluated
in light of their explanatory power. One potential dimension that could allow
the variation necessary among distinct cognitive mechanisms is environmental
variation. Controlled laboratory conditions often mask the inter and
intra-specific variation that arises in the interaction between mechanisms and
environments. Thus, if predictions from the alternative mechanisms change
with environmental conditions, it is possible to asses the explanatory power
of the different hypothesis by performing behavioural test in the context of
different environmental conditions.

From a evolutionary perspective, mechanisms are likely selected because
of how they allow individuals to respond to environmental variation.
For example, biological market theory predicts that the exchange rate of goods
and/or services traded between cooperative partners adjusts to the
law of supply and demand, when individuals have some degree of partner choice
\citep{noe_Biological_1995}. Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should favour the ability
to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent
mechanisms beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Client fish seek cleaner fish services at their territory
(so-called ``cleaning station'') and offer themselves as food patches
to get their ectoparasites removed, which provides cleaners
with food and clients with improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016, demaire_Reduced_2020}.
Given the capacity of some client fish to swim larger distances and
access multiple cleaning stations while others access the only cleaning
station at their territory, it is crucial to categorize clients as
either ``visitors'' or ``residents'', respectively. During cleaning interactions,
a cleaner fish often face a choice between a visitor and a resident client
seeking its cleaning services simultaneously. Visitors have the option
to switch to another cleaner fish if being made to wait, while residents
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001}.

When researchers aimed at testing wild-caught cleaner fish, as
well as individuals from other species, abilities
to prioritize visitor over resident clients in a lab-based paradigm,
they used Plexiglas plates of different colours and shapes offering same
amount of food as surrogates for visitor and resident clients.
One plate is made to behave like a resident, i.e.~it remains until
the cleaner fish had eaten off it. The other plate was made to behave
like a visitor, i.e.~it is removed if not inspected first.
This laboratory paradigm originally inspired by the cleaning market
of cleaner fish is more generally refered as the ephemeral reward task.
Individuals are assesed in their capacity to develop a preference for the
ephemeral option, which gives them access to two reward units.
Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Furthermore, adult cleaner fish outperformed various primates as well as rats and
pigeons in this original version of the ephemeral reward task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; yet the African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. This interspecific
variation in performance, and primates' known cognitive abilities \(\color{red}{\text{ref}}\), shows that performane in the task
cannot be readily explained by classifying species as having high and
low cognitive abilities. Instead, this inter-spefic variation suggest
that solving the ephemeral reward task requires a set of cognitive
mechanisms that are not present in all the tested taxa, or were not used
under the experimental set-up.

@@

To solve the ephemeral reward task, cleaners need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is in principle
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, however, the two
mechanisms should be evaluated in light of how well
they explain the available data. Note that different fields interested in
cognition and desicion making use different words to refer to negative
reinforcers \citep{quinones_Reinforcement_2019, sutton_Reinforcement_2018}.
In here, for the sake of simplicity and clarity, we will use the word
`penalty' to refer to this mechanism which includes a negative reinforcer.

@@

The interplay between the underling cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether chaining, penalty, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show low preference for visitors under low
local cleaner-to-client ratio.

@@

The model shows that agents need to find a mean
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates.
If \(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action (chaining).
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor as a negative component (penalty). Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

@@

The cognitive performance data was from a total of 120 cleaners
(10 individuals per 12 site/year) tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner fish inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. In the original studies, once a fish reach a learning criterion,
that is, performing significantly above chance level (\(> 50%
\), \(p-value ≥ 0.05\)),
they passed to a reversal version of the task where the roles of the
visitor/resident Plexiglas plates were swapped (see Triki et al.~2019, 2020).
Here, we used instead a subset of these data in order to have an idea on
cleaner fish preferences for the visitor plate, even if they do not reach
the learning criterion within 200 trials. To do so, we first extracted the
trial-by-trial outcomes from the last two session (20 trials) of those who
never reached the learning criterion for visitor plate
(\(N = 45\) cleaner fish).
For those who reached the learning criterion at some point during the
test and passed to a reversal phase, we extracted the trials-by-trial
outcomes from the last session (10 trials) before passing to reversal
and the last 10 trials they were exposed to in the test
(\(N = 75\) cleaner fish).
We chose a combination of initial and reversal to quantify preference
for the visitor client, given that it matches well the criteria chosen
in previous analysis of the ephemeral reward task experimental set-up
\citep{triki_Biological_2019, triki_Brain_2020}. In the
suplementaty material (Fig. \ref{fig:rawdata}) we show how our choice maps
to the previously used criteria. For comparison, we also show a quantification
of visitor preference based only on the initial phase of the experiments.

@@

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/post-1}

\caption{Posterior distributions for parameter values \(\gamma\) and \(\eta\) for
the three models (``Full model'', ``chaining only'' and ``penalty only''). We show
the kernel density estimates, below the mode (black dot) and the 65\%
(light blue shade) and 95\% (grey shade) highest posterior density interval
for the two parameters. On the top, panels a and b show posterior
distributions for a full model, including chaining (\(\gamma\)) and penalty
(\(\eta\)). Panel c, shows the \(\gamma\) estimate from model with only chaining.
Panel d shows the \(\\eta\) estimate from a model with only penalty. Panel e,
shows a measure of fit for all models, namely the distribution of pseudo-\(R^2\)
obtained from sampling the posterior distribution of parameter values.}\label{fig:post}
\end{figure}

Estimation of parameter values for the three models (full model,
chaining and penalty), support chaining as the only mechanism
cleaners use to account for the future effects of their actions;
and thus to solve the ephemeral reward task. In the estimation of the
parameter values of the full model, which includes both chaining and
penalty, the bulk of the marginal posterior distribution of \(\eta\) which
controls the strength of penalty is around 0 (Fig. \ref{fig:post}).
As for \(\gamma\), controlling chaining, the 95\% confidence
intervals also includes zero, but the mode of the posterior
is around 0.5 (Fig. \ref{fig:post}, a). In the chaining model,
where \(\eta\) is set to zero, the distribution of \(\gamma\) shifted
to higher values, zero is no longer part of the 95\% credible interval
of the parameter (Fig. \ref{fig:post}, c). In contrast, when we
look at the model with only penalty, the posterior distribution
of \(\eta\) is still centred around zero (Fig. \ref{fig:post}, d).
Thus, the analysis of the estimates of individual parameter
values in the three models only supports a strong
effect of chaining. Furthermore, the comparison of the models` fit
favours the chaining model. In panel e of figure \ref{fig:post}
we show the distribution of \(pseudo-R^2\) calculated using
samples from the posterior distributions shown before.
Note, \(pseudo-R^2\) can have negative values, which is when
the log-likelihood of the model is lower than that of a model that
triggers neutral preferences. Even though, the peak of the three
\(pseudo-R^2\) distributions was not very different, the model with only
chaining produced a distribution of \(pseudo-R^2\) where more values
were positive (to the right of black line in Fig. \ref{fig:post} f). This
shows that, accounting for variation in the parameter estimates, the
model with chaining gives a better fit to the data, despite
having one parameter less than the full model. We have not shown
here the marginal posterior distributions of the scaling constant,
given that they do not bring biological insight.
Their visualizations can be found in the supplementary
material (Fig. \ref{fig:scaConst}), as well as the diagnostics of the MCMCs
(Figs. \ref{fig:diagnosticsfull},\ref{fig:diaggam},\ref{fig:diagNeg}).

{\centering \includegraphics[width=1\linewidth]{manuscript_1.0_files/figure-latex/pred-1}

\caption{Observed and predicted probability of choosing a visitor.
Left-hand side panel: colour contour shows the prediction from the
learning model using the mode of the posterior distributions of parameters
recovered by the statistical analysis. Dots show the frequency of visitor
choices for the 12 reef sites, as well as the corresponding relative
cleaner fish abundance (x axis) and frequency of visitors clients leaving
the cleaning station (y axis). Right-hand side panels: Variation of the
predicted probabilities of choosing a visitor over a resident and their
observed values for 12 locations. Circles show the mean prediction for
each location from 100 samples taken from the posterior distribution.
Thick and thin bars show the 66 and 95\% credible interval, respectively,
taken from those posterior samples. Squares show the predictions used for
panel on the right-hand side. Colour coding denotes different reef sites/year
of the data collection (see Empirical data section). Black line corresponds
to a perfect match between observed and predicted probabilities. Upper
panels (a and b) show predictions from a model including chaining and
penalty; middle panels (c and d) from a model with only chaining; lower
panel (e and f) from a model with only penalty.}\label{fig:pred}
\end{figure}

The main reason for chaining and penalty to give different
predictions is the way that cleaner fish relative abundance influences the
preference for the visitor clients. Visitor leaving probability has a
similar positive effect on the probability of choosing the visitor
clients on all three models, they all predict an increase preference
for visitors as the visitor probability increases (Fig. \ref{fig:pred}).
In contrast, cleaner fish relative abundance has a different effect in the
model with only chaining, compared to the other two models,
full model and penalty. In the model with only chaining, only intermediate
cleaner fish abundance triggers a preference for the visitor
clients (Fig. \ref{fig:pred} c). In both models with penalty,
both intermediate and low cleaner abundance triggered a preference
for the visitor (Fig. \ref{fig:pred} a,e). Note, however, we calculated
preferences shown in figure \ref{fig:pred} left panels by using only the mode
of the posterior distributions, and by holding constant the balance between
resident and visitors´s abundances. Panels on the right, show how close
predictions are from the observed data, allowing the balance between
client types to vary, and using a set of samples from the posterior
distribution.

@@

Previous research showed that cleaner fish living at high population
densities and giving service priority to the visitor plate in the
ephemeral reward task, as well as cleaner fish living at low
densities but denying service priority to the visitor plate possess
larger forebrains; a key teleost brain region associated with
behavioural flexibility and social intelligence. Those failing to
show optimized decision-rules given their local ecological
conditions had relatively smaller forebrains \citep{triki_Brain_2020}.
Triki \emph{et al.} refer to the former as socially competent cleaner fish,
while the second group as socially incompetent cleaner fish.
Social competence is the ability to optimise social behaviour
to the available social information \citep{taborsky_Social_2012, bshary_Cooperation_2015, varela_Correlated_2020}. Our analyses yielded no
evidence that the difference in social competence with respect to the local
ecological conditions and associated brain morphology,
found by Triki \emph{et al} \citep{triki_Brain_2020}, are due to the mechanism
used to incorporate future consequences. It is conceivable that
high performing individuals from low population densities reef sites
use negative reinforcement instead of chaining, but in that
case negative reinforcement should have explained at least
part of the data. Configurational learning or
chunking \citep{sutherland_Configural_1989, miller_Magical_1956},
the second component necessary to solve the ephemeral
reward task \citep{quinones_Reinforcement_2019}, was
not varied in the models we analized here. However, while chunking tendencies
should vary to allow individuals to adapt to their local conditions
\citep{prat_Modelling_2022, kolodny_Evolution_2014}, systematic differences
in individual chunking tendencies would not explain how socially
competent decisions vary as a function of relative abundance.
Therefore, it remains currently unclear what cell-demanding
mechanisms may cause variation in social competence that
translates into site-specific variation in performance in the
ephemeral reward task.

Our models are inspired by the general processes of associative learning
where short term rewards are translated into decision making; thus, it
ignores alternative channels of information that could be relevant in
market-like situations. For example, the model does not investigate whether
cleaner fish actually assess the frequency of client visits or a mean frequency
of visitors leaving. The updating learning mechanism for the development
of preferences works on a trial-by-trial basis. In the model, cleaner fish
do not need to assess the actual state of the market, \emph{i.e.} their abundance,
the abundance of residents and visitors, and client visitation rate as an
indicator of demand. They only need to assess short-term consequences of their
own decisions on food intake, and chain them. Also, for the sake of
simplicity, the model ignores the process by which cleaner fish discriminate
residents and visitor clients. A model that accounts for this discrimination
probably would involve the development of preferences for morphological or
behavioural features that are statistically associated with visitors
or residents. For example, visitors are on average larger
than residents in body size \citep{bshary_Cleaner_2001}, and contrary to
resindents, are less likely to chase a cleaner fish that fails to cooperate
and intead cheats its client by taking a bite of mucus
\citep{bshary_Asymmetric_2002}. Given these associations, chaining might
produce the decision-rule ``choose the larger client and/or the less aggressive
client'', which is not a useful rule in the standard ephemeral reward task.

In conclusion, our study shows that variation in cognitive performance as
a function of the local ecological conditions may set the stage for the use of
mechanistic modelling to identify the cognitive processes underlying
learning in animals. The combination overcomes the limitations
of the general philosophy in animal cognition to apply the logic of
Lloyd Morgen's canon (Occam's razor). Cognitive experiments with the
aim of excluding basic reinforcement learning as a potential
explanation (operant and/or classical conditioning) of performance
often employ one trial experiments requiring animals to solve the
task on the first possible occasion.
For example, any theory of mind task needs to be solved in the first
trial in order to exclude fast conditioning \citep{heyes_Theory_1998}.
Similarly, subjects need to solve a social learning task on the first
trial to accept imitation as a mechanism over stimulus/local
enhancement. Such strict conditions are virtually never met. For
example, potato washing by Japanese macaques, an iconic example of
social learning, took several years to spread within the group
\citep{kawamura_Process_1959}, meaning that any learner had been repeatedly
exposed to demonstrations before acquisition. Importantly, Galef
\citep{galef_Question_1992} refuted imitation as mechanism not simply because
of the repeated exposure but because a (rather qualitative) analysis of
the spread of potato washing across individuals did not follow the
prediction based on imitation learning (see also
\citep{hirata_SweetPotato_2001}). In our case, the number of trials it took
cleaners to learn the solution to the ephemeral rewards task
would never allow to exclude an important role of
penalty based on the data
alone. However, fitting model predictions to our comprehensive empirical
data set revealed that a more complex mechanism, estimation of future reward,
fits the data better.
